Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-29 paper.tex
Move recombination equation into main text.

Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
and empirical recombination maps similar to other
simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and results from \ms.

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination
model~\citep{mcvean2005approximating,marjoram2006fast},
and was originally motivated to a significant degree by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes and several approaches have been proposed to
allow simulations that more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
The SMC is an important analytical approximation
and has allowed the development of numerous inference
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust}.

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Approximate run time for \msprime\ simulations,
    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black line is a quadratic fit separately in each panel.
}
\end{figure}

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
Estimates of the number of such ``trapped'' recombinations grow quadratically
with genome length \citep[Eq.~5.10]{hein2004gene}.
Because they do not track such recombinations, efficient SMC implementations
will be faster than \msprime\ for sufficiently high recombination rate,
genome length, or $N_e$. %Until we decide to approximate the recombination process
%between distant ancestral segment. Hint hint.

% \jkcomment{We want to round out this discussion here by giving some
% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
% to Fig.~\ref{fig-ancestry-perf} to show roughly what we can expect
% to simulate in how long. See issue 28.}

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
(\citet{hein2004gene} gives an equation
including the effects of sample size --
see Appendix~\ref{app-theoretical-runtime}.)
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plot show, roughly, that simulating
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.

However, what about a simulation with \emph{changing} size?
To understand how runtime depends on demography
it helps to consider \emph{why} runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

@@

The SMC~\citep{mcvean2005approximating} and
SMC'~\citep{marjoram2006fast} models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore only appropriate for studying the SMC approximations
themselves; an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
simulation of organisms such as \textit{Drosophila melanogaster}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=7cm]{figures/gc-perf}
\end{center}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

\section*{Theoretical runtime}
    \label{app-theoretical-runtime}

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf-expected}
\end{center}
\caption{\label{fig-expected-perf}
    Data from Figure~\ref{fig-ancestry-perf}, compared with
    equation~\eqref{eqn-hsw-bound} (``expected'').
    Left panel shows observed times plotted against expected number
    of operations; right panel shows the ratio of the two, plotted against $\rho$.
}
\end{figure}

The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
on the number of events involving trapped ancestral material in Hudson's algorithm.
For $n$ samples from
a diploid population of effective size $N_e$,
and a genome of genetic length $r$,
with $\rho = 4 N_e r$ the scaled recombination rate,
and $R^T$ the number of recombination recombination events in trapped material,
this upper bound is
\begin{equation} \label{eqn-hsw-bound}
    \mathbb{E}[R^T]
    \le
    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
\end{equation}
Note that the sum on the right is well-approximated by $\log(n)$.
Although \citet{hein2004gene} refer to this bound as ``crude'',
it turns out to predict the scaling of runtime quite well for large genomes.
Figure~\ref{fig-expected-perf} shows the observed runtime
(from the same simulations as Figure~\ref{fig-ancestry-perf})
plotted against equation~\eqref{eqn-hsw-bound},
showing that for values of $\rho$ above $4 \times 10^4$ or so
(i.e., chromosome-length simulations in populations above $10^4$),
the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
This implies the runtime for a large sample
can be predicted from the runtime for a small sample.
For smaller values of $\rho$, however, the prediction
severely underestimates runtime.

>>>>

Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots and patterns
of recombination from empirical maps.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and \ms.

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination
model~\citep{mcvean2005approximating,marjoram2006fast},
and was originally largely motivated by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes and several approaches have been proposed
to allow simulations more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
The SMC and SMC' models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore currently only appropriate for studying the
SMC approximations themselves, although we hope to
implement them more efficiently in future versions.

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
    \includegraphics{figures/ancestry-perf-expected}
\end{center}
\caption{\label{fig-ancestry-perf}
    Approximate run time for \msprime\ simulations,
    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black line is a quadratic fit separately in each panel.
}
\end{figure}

% \begin{figure}
% \begin{center}
% \end{center}
% \caption{\label{fig-expected-perf}
%     Data from Figure~\ref{fig-ancestry-perf}, compared with
%     equation~\eqref{eqn-hsw-bound} (``expected'').
%     Left panel shows observed times plotted against expected number
%     of operations; right panel shows the ratio of the two, plotted against $\rho$.
% }
% \end{figure}

In human-like parameter regimes and for large sample sizes,
\msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments
and for large $\rho$, these events will dominate.
\citet[Eq.~5.10]{hein2004gene} derived an upper bound
on $R_T$, the number of recombination events within trapped ancestral material
in Hudson's algorithm,
\begin{equation} \label{eqn-hsw-bound}
    \mathbb{E}[R_T]
    \le
    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
\end{equation}
Therefore, since the sum on the right is well approximated by $\log{n}$,
we can very roughly say that the time
complexity of Hudson's algorithm is $O(\rho^2 \log^2 n)$.

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plot show, roughly, that simulating
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.

Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
as ``crude'', it predicts the scaling of run time quite well for large genomes
Figure~\ref{fig-expected-perf} shows the observed runtime
(from the same simulations as Figure~\ref{fig-ancestry-perf})
plotted against equation~\eqref{eqn-hsw-bound},
showing that for values of $\rho$ above $4 \times 10^4$ or so
(i.e., chromosome-length simulations in populations above $10^4$),
the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
This implies the runtime for a large sample
can be predicted from the runtime for a small sample.
For smaller values of $\rho$, however, the prediction
severely underestimates runtime.

However, what about a simulation with changing population size?
To understand how runtime depends on demography
it helps to consider why runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

@@

an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
simulation of organisms such as \textit{Drosophila melanogaster}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=7cm]{figures/gc-perf}
\end{center}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 Mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

% \section*{Theoretical runtime}
%     \label{app-theoretical-runtime}
% \begin{figure}
% \begin{center}
%     \includegraphics{figures/ancestry-perf-expected}
% \end{center}
% \caption{\label{fig-expected-perf}
%     Data from Figure~\ref{fig-ancestry-perf}, compared with
%     equation~\eqref{eqn-hsw-bound} (``expected'').
%     Left panel shows observed times plotted against expected number
%     of operations; right panel shows the ratio of the two, plotted against $\rho$.
% }
% \end{figure}

% The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
% on the number of events involving trapped ancestral material in Hudson's algorithm.
% For $n$ samples from
% a diploid population of effective size $N_e$,
% and a genome of genetic length $r$,
% with $\rho = 4 N_e r$ the scaled recombination rate,
% and $R^T$ the number of recombination recombination events in trapped material,
% this upper bound is
% \begin{equation} \label{eqn-hsw-bound}
%     \mathbb{E}[R^T]
%     \le
%     \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
% \end{equation}
% Note that the sum on the right is well-approximated by $\log(n)$.
% Although \citet{hein2004gene} refer to this bound as ``crude'',
% it turns out to predict the scaling of runtime quite well for large genomes.
% Figure~\ref{fig-expected-perf} shows the observed runtime
% (from the same simulations as Figure~\ref{fig-ancestry-perf})
% plotted against equation~\eqref{eqn-hsw-bound},
% showing that for values of $\rho$ above $4 \times 10^4$ or so
% (i.e., chromosome-length simulations in populations above $10^4$),
% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
% This implies the runtime for a large sample
% can be predicted from the runtime for a small sample.
% For smaller values of $\rho$, however, the prediction
% severely underestimates runtime.
